Building an SSI Cluter with GNU/Hurd

### 2019-05-26 libpager/paper.tex
working on "Building an SSI Cluster" paper

\usepackage{nopageno}
\usepackage{vmargin}
\usepackage{comment}
\usepackage{xspace}
\usepackage{hyperref}
\usepackage{mdframed}
\usepackage{xcolor}

@@

% Shortcuts just for this document
\def\libpager{{\tt libpager}\xspace}
\def\netmsg{{\tt netmsg}\xspace}
\def\pagersync{{\tt pager_sync()}\xspace}
\def\pagerwritepage{{\tt pager_write_page()}\xspace}
\def\pagerreadpage{{\tt pager_read_page()}\xspace}
\def\moinit{{\tt memory_object_init}\xspace}
\def\modatasupply{{\tt memory_object_data_supply}\xspace}
\def\modataerror{{\tt memory_object_data_error}\xspace}

@@

@@

\tikzset{
    state/.style={
        draw, ellipse
    },
    >=latex
}

\def\notm{$\overline{\rm m}$}
\begin{tikzpicture}[node distance=3cm]

@@

\draw [->] (e) -- node[midway]{{\bf data_request}} node[midway, below]{(read)} (wait-r);
\draw [->] (e) -- node[midway]{{\bf data_request}} node[midway, below]{(write)} (wait-w);
\draw [->] (wait-r) -- (wait-w);
\draw [->] (wait-r) -- node[midway]{{\bf data_supply}} node[midway, below]{(no access)} (nr);
\draw [->] (wait-r) -- node[midway]{{\bf data_supply}} node[midway, below]{(read access)} (rr);
\draw [->] (wait-r) to[out=45, in=135] node[midway, above]{{\bf data_supply}} node[midway, below]{(write access)} (ww);

\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(no access)} (nw);
\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(read access)} (rw);
\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(write access)} (ww);

\draw [->] (wait-r) to[out=-135, in=-135] node[midway]{{\bf data_error}} (e);
\draw [->] (wait-w) to[out=-45, in=-45] node[midway]{{\bf data_error}} (e);

\end{tikzpicture}
\end{center}
\caption{Page-in transitions.  Adapted from figure 6 in [1]}
\label{Mach memory}
\end{figure}

Each virtual memory page in the Mach kernel can be in one of fifteen
different states: empty (e), waiting for either read or write access,
and twelve different states for pages in memory, depending on
whether the page has been modified or not, the current access, and the
desired access.  Figure \ref{Mach memory} shows the transitions
involved in paging in; [1] uses five similar diagrams to completely
illustrate all of the Mach memory manager's possible transitions.

Mach's ability to attach out-of-band memory to messages is thus a
crucial component in the efficient implementation of its memory
management.  A page is typically supplied to kernel by attaching it to
a \modatasupply message.  No copying needs be done, as the
page is transferred to kernel control simply by adjusting bits in the
hardware memory management tables.  Likewise, transferring a modified
page from the kernel to a user-space memory manager is done by
attaching it to a Mach message from the kernel.

@@

I wrote Hurd's current \netmsg server in 2016.  It operates over a
single unsecured TCP/IP connection and has many limitations, but works
well enough to implement distributed shared virtual memory.  In server
mode, it presents a single node, by default the root of the
filesystem, to any clients that connect.  In client mode, it acts as a
translator, allowing a remote server's exported node to be mounted in
the local filesystem.

EXAMPLE: a \netmsg session

Thus, we have a remote file system, but it's important to realize that
this remote file system supports full Mach messaging.  Opening a file,
for example, is done by sending a Mach io_? message that is relayed
across the network, then received, processed, and answered by the
remote Hurd server managing that file.  This is in contrast to remote
file systems like NFS that don't transport Mach messages, and only
allow a subset of Hurd RPCs which can translated into NFS messages.
In particular, since Mach virtual memory is implemented using the Mach
messaging protocol, \netmsg allows a remote file to be {\tt mmap}'ed
into local memory.

\netmsg works by copying Mach messages almost verbatim across the
network connection.  Only the Mach port numbers are changed.
Out-of-band memory is simply copied across the network connection,
then deallocated.

While usable, this version of \netmsg has many flaws.  Transporting
all of the Mach messages over a single TCP/IP session provides
reliable, in-order delivery as guaranteed by Mach, but at the cost of
serializing all messages between two hosts.  Also, using the system's
TCP/IP stack requires context switches between \netmsg and the
TCP/IP server (a user-space program, like so much else in Hurd) for
every Mach message being relayed.  The current scheme doesn't do a
good job of handling more than two nodes, because there's no mechanism
to detect a cycle, which could result in a lot of unnecessary network
chatter.  Finally, it's become clear that the simplistic means of
managing ports creates unavoidable race conditions.

For example, consider (again) what happens when node A drops its last
send right to a port on node B.  Node A sends a NO SENDERS message to
node B and forgets about the port.  It's possible that node B has sent
node A another send right for the same port, and that it's in transit
when node A sends the NO SENDERS message.  In that case, node A would
actually have an outstanding send right when node B receives the NO
SENDERS message, which would cause node B to incorrectly drop the send
right linked to node A.  I can see no way around this except a more
complicated handshake where the NO SENDERS message from node A to node
B would include a sequence number allowing node B to determine that it
hadn't sent any additional send rights, and a reply message from node
B to node A finalizing the port deallocation.  (what about SEQNOS?)

@@

Hurd's implementation of the Mach memory management protocol is
implemented in a library called ``libpager''.  The original \libpager,
written in the mid-1990s, only supported a single client (i.e, a
single Mach kernel), and was thus suitable for single node operation.
Its API, however, is flexible enough to allow a more sophisticated,
multi-client \libpager as a drop-in replacement.  For dynamically
linked servers, all that is required is to update the system shared
library to obtain multi-node functionality.

@@

How do we figure out which client is the local kernel?  One possible
solution is for \netmsg to listen on /servers/netmsg for an RPC that
accepts a send right and identifies which node currently hosts the
corresponding receive right.  Mach 3 Kernel Interfaces defined the
{\tt norma_port_location_hint} RPC that might be useful.

@@

Two bits were allocated to each of two error codes.  {\tt ENOSPC} and
{\tt EDQUOT} were preserved; all other errors get collapsed into {\tt
  EIO}.  The remaining bits indicate if the page is in core, if we're
waiting for it to be paged in or out to the backing store.

@@

Handling error on unlock requests is somewhat tricky.  {\it Mach 3
  Kernel Interfaces} states, in its {\bf NOTES} on {\bf
  memory_object_lock_request}:

@@

\subsection{Pseudocode}

@@

\section{POSIX Shared Memory}

Hurd's System V shared memory implementation, in glibc's
sysdeps/mach/hurd, operates by mapping "files" in the /run/shm portion
of the filesystem tree.  /run/shm is typically implemented using the
tmpfs translator, which is linked with \libpager.  Thus, a multi-client
\libpager enables the use of inter-node shared memory.  What is
required to make this happen is a shared /run/shm - shared with full
Mach IPC semantics, not just NFS - so \netmsg is required.  unionfs can
possibly be used to overlay the various /run/shm's on each other.

\section{Distributed Tasks and Process Migration}

In the Hurd architecture, processes are localized to a node, but our
implementation of distributed shared memory allows cross-node forks
with full POSIX semantics.  A straightforward modification to
Hurd's C library would allow fork calls to spawn processes
on other nodes.

How can a process share threads between nodes?  Two main problems:
sharing Mach ports, and sharing memory.  Mach allows the allocation
of un-backed memory pages.  We need to attach a memory manager
to the un-backed pages to allow them to be distributed across nodes.
This requires relatively minor changes to Mach.

This would allow cross-node forks, but not cross-node threads, which
would be difficult to implement using Mach's current design.  Any
thread in a task can listen for messages on a Mach port, and messages
are only delivered once, so balancing the need to copy messages to
multiple nodes (since any thread could be the only one listening to a
given port) while guaranteeing unique delivery is quite problematic.

This suggests a natural division: processes are local to each node.  A
multi-threaded program can fork multiple times, and each forked
process can spawn multiple threads.  The processes are distributed
between nodes, while each process is local to a node and supports
multiple threads.

Linux's NUMA implementation uses a zonelist for memory allocation.
Memory allocation requests are attempted locally, and if this fails,
then we fall back into the zonelist, which gives us a list of other
nodes to try for memory allocation.

\section{Blocking Behavior of {\tt mach_msg()}}

An issue with Mach's design caused me no little bit of grief.
Mach's {\tt mach_msg()} (as I read the documentation),
allows for non-blocking sends: according to {\it Mach 3
  Kernel Interfaces}:

\item Useful Mach kernel changes

>>>>

\usepackage{nopageno}
\usepackage{vmargin}
\usepackage{comment}
\usepackage{xspace}
\usepackage[hyperfootnotes=false]{hyperref}
\usepackage{mdframed}
\usepackage{xcolor}

@@

% Shortcuts just for this document
\def\libpager{{\tt libpager}\xspace}
\def\netmsg{{\tt netmsg}\xspace}
\def\pagersync{{\tt pager_sync()}\xspace}
\def\pagerwritepage{{\tt pager_write_page()}\xspace}
\def\pagerreadpage{{\tt pager_read_page()}\xspace}
\def\moinit{{\tt memory_object_init}\xspace}
\def\moready{{\tt memory_object_ready}\xspace}
\def\morequest{{\tt memory_object_data_request}\xspace}
\def\mosupply{{\tt memory_object_data_supply}\xspace}
\def\modataerror{{\tt memory_object_data_error}\xspace}

%% Disable hypertext links in table of contents
%% https://tex.stackexchange.com/questions/55190

\makeatletter
\let\Hy@linktoc\Hy@linktoc@none
\makeatother

@@

\tableofcontents

\vfill\eject

@@

\tikzset{
    state/.style={
        draw, ellipse
    },
    trans/.style={
        line width=2pt,
        transparent!20,
        ->,
        >=latex
    },
    label/.style={
      midway, align=center,
      transparent!100
    }
}

\def\notm{$\overline{\rm m}$}

\begin{tikzpicture}[node distance=3cm]

@@

\draw [trans] (e) -- node[label] {{\bf data_request}\\(read)} (wait-r);
\draw [trans] (e) -- node[label] {{\bf data_request}\\(write)} (wait-w);
\draw [trans] (wait-r) -- (wait-w);
\draw [trans] (wait-r) -- node[label] {{\bf data_supply}\\(no access)} (nr);
\draw [trans] (wait-r) -- node[label]{{\bf data_supply}\\(read access)} (rr);
\draw [trans] (wait-r) to[out=45, in=135] node[label]{{\bf data_supply}\\(write access)} (ww);

\draw [trans] (wait-w) -- node[label]{{\bf data_supply}\\(no access)} (nw);
\draw [trans] (wait-w) -- node[label]{{\bf data_supply}\\(read access)} (rw);
\draw [trans] (wait-w) -- node[label]{{\bf data_supply}\\(write access)} (ww);

\draw [trans] (wait-r) to[out=-135, in=-135] node[label]{{\bf data_error}} (e);
\draw [trans] (wait-w) to[out=-45, in=-45] node[label]{{\bf data_error}} (e);

\end{tikzpicture}
\end{center}
\caption{Page-in transitions.  Adapted from Figure 6 in \cite{principles}}
\label{Mach memory}
\end{figure}

Each virtual memory page in the Mach kernel can be in one of fifteen
different states: empty, waiting for either read or write access, and
twelve different states for pages in memory, depending on whether the
page has been modified or not, the current access, and the desired
access.  Figure \ref{Mach memory} shows the transitions involved in
paging in; \cite{principles} uses five similar diagrams to completely
illustrate all of the Mach memory manager's possible transitions.

Mach's ability to attach out-of-band memory to messages is thus a
crucial component in the efficient implementation of its memory
management.  A page is typically supplied to kernel by attaching it to
a \mosupply message.  No copying needs be done, as the
page is transferred to kernel control simply by adjusting bits in the
hardware memory management tables.  Likewise, transferring a modified
page from the kernel to a user-space memory manager is done by
attaching it to a Mach message from the kernel.

For example, consider how Hurd handles an {\tt mmap()} call.  A
program uses directory lookup calls to obtain a file port, which is a
send right to a port managed by a file server, and corresponds to a
standard POSIX file descriptor.  It then uses the {\tt io_map} RPC to
obtain another send right, also to a port managed by the file server,
which implements the Mach memory management protocol.  It passes that
send right to the kernel, using a {\tt vm_map} RPC, indicating that it
wishes to map that memory object (i.e, the file), into its memory
space.  The kernel, after a \moinit / \moready handshake with the file
server, creates a new memory region in the program's address space,
initially with no physical memory assigned to it, and returns that
address to the program.

As soon as the program attempts to access this region of memory in any
way, a page fault is triggered, which causes the kernel to send an
\morequest message to the file server.  The file server responds by
reading the required page from disk (if necessary) and passing it to
the kernel by attaching it to a \mosupply message.  The kernel, now
managing this region of memory, maps it to the correct page in the
previously assigned region of memory, and allows the program to
continue execution.

If the file server had instead replied with \modataerror, the program
would have received a seg fault instead.  If another program maps the
same file, the kernel will not request duplicate copies of the pages,
but will arrange for the physical memory to be shared between the
processes.  Also, although the kernel now manages the memory, the file
server can request it back at any time; the model is that the file
server manages the memory, and the kernel is only caching it.  If the
kernel runs short on physical memory, it can either discard the page,
if it is unmodified and if the file server didn't set the ``precious''
flag, or will return it to the file server.

\tikzstyle{kernel} = [draw, fill=blue!20, rounded corners, minimum width=30em]

\begin{center}
\begin{tikzpicture}

\task{Filesystem Server}{at (0,0)}{}{
%\node (memory1) [above left=-0.5, fill=white, minimum size=1.5cm, dashed, thick, draw] {};
}
\task{Program}{at (15 em,0)}{}{
%\node (memory1) [above right=-0.5, fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
\node (memory2) [above right=-0.5, fill=white, minimum size=1.5cm, dashed, thick, draw] {};
\node [below] at (memory2.north) {\tiny addr};
}

\node (kernel) [kernel] at (7.5 em, -7 em) {Kernel};
%%\node (kmemory) [fill=white, minimum size=1.5cm, dashed, thick, draw] at (7.5 em, -10 em) {};
%%\node [below] at (kernel.north) {Kernel};

\draw [color=red, <-, thick] (2 em,1) -- node [midway, above] {io\_map} (13 em,1);
\draw [color=red, ->, thick, dotted] (2 em,.9) -- node [midway, below] {\tiny port 153} (13 em,.9);

\draw [color=red, ->, thick] (15 em,-3 em) -- node [midway, left, align=center] {vm\_map\\{\tiny port 153}} (15 em,- 7 em);
\draw [color=red, <-, thick, dotted] (15.2 em,-3 em) -- node [midway, right] {\tiny addr} (15.2 em,- 7 em);

\draw [color=red, <-, thick] (-2 em,-3 em) -- node [midway, left] {m\_o\_init} (-2 em,- 7 em);
\draw [color=red, ->, thick] (-1.5 em,-3 em) -- node [midway, right] {m\_o\_ready} (-1.5 em,- 7 em);

%\draw [color=red, <-, thick] (0.5 em,-3 em) -- node [midway, left] {m\_o\_data\_request} (0.5 em,- 8 em);
%  \draw [color=red, ->, thick] (1 em,-3 em) -- node [midway, right] {m\_o\_data\_supply} (1 em,- 8 em);

\end{tikzpicture}
\end{center}

@@

I wrote Hurd's current \netmsg server in 2016.  It operates over a
single unsecured TCP/IP connection and has many limitations, but works
well enough to implement distributed shared virtual memory.  In server
mode, it presents a single node, by default the root of the
filesystem, to any clients that connect.  In client mode, it acts as a
translator, allowing a remote server's exported node to be mounted in
the local filesystem.
\netmsg works by copying Mach messages almost verbatim across the
network connection.  Only the Mach port numbers are changed.
Out-of-band memory is simply copied across the network connection,
then deallocated.

\begin{figure}[h]
\begin{mdframed}
\begin{verbatim}
root@qemu-hurd-1:~/github/netmsg# ./netmsg -s ~ &
[1] 761
root@qemu-hurd-1:~/github/netmsg# 
\end{verbatim}
\end{mdframed}

\begin{mdframed}
\begin{verbatim}
root@qemu-hurd-2:~/netmsg/netmsg# settrans -ac node ./netmsg 192.168.55.101
root@qemu-hurd-2:~/netmsg/netmsg# ls node
github
hurd-0.9.git20190331
hurd_0.9.git20190331-5.debian.tar.bz2
hurd_0.9.git20190331-5.dsc
hurd_0.9.git20190331.orig-eth-filter.tar.bz2
hurd_0.9.git20190331.orig-libddekit.tar.bz2
hurd_0.9.git20190331.orig-libdde-linux26.tar.bz2
hurd_0.9.git20190331.orig-libmachdev.tar.bz2
hurd_0.9.git20190331.orig.tar.bz2
old
translator_primer
root@qemu-hurd-2:~/netmsg/netmsg# 
\end{verbatim}
\end{mdframed}
\caption{A \netmsg session}
\label{A netmsg session}
\end{figure}

For example, Figure \ref{A netmsg session} shows a simple \netmsg
session between two Hurd virtual machines.  On {\tt qemu-hurd-1}, I
started a \netmsg server ({\tt -s}) sharing {\tt root}'s home
directory, listening on \netmsg's default TCP port 2345.  Then, on
{\tt qemu-hurd-2}, I attached an active ({\tt -a}) translator to {\tt
  node}, creating it ({\tt -c}) if it didn't already exist, running
\netmsg as a client connecting to {\tt qemu-hurd-1}'s IP address.
Running {\tt ls node} shows the remote directory.

Thus, we have a remote file system, but it's important to realize that
this remote file system supports full Mach messaging.  Reading a file,
for example, is done by sending a Hurd {\tt io_read} message that is
relayed across the network, then received, processed, and answered by
the remote Hurd server managing that file.  This is in contrast to
remote file systems like NFS that don't transport Mach messages, and
only allow a subset of Hurd RPCs which can translated into NFS
messages.  In particular, since Mach virtual memory is implemented
using the Mach messaging protocol, \netmsg allows a remote file to be
{\tt mmap}'ed into local memory.

While usable, this version of \netmsg has many flaws.  Transporting
all of the Mach messages over a single TCP/IP session provides
reliable, in-order delivery as guaranteed by Mach, but at the cost of
serializing all messages, and pausing all communication to wait for a
TCP re-transmission, when only individual Mach ports need to be
serialized.  Also, using the system's TCP/IP stack requires context
switches between \netmsg and the TCP/IP server (a user-space program,
like so much else in Hurd) for every Mach message being relayed.  The
current scheme doesn't do a good job of handling more than two nodes,
because there's no mechanism to detect a cycle, which could result in
a lot of unnecessary network chatter.  Finally, it's become clear that
the simplistic means of managing ports creates unavoidable race
conditions.

% However, once I had a minimally working \netmsg, I moved
% on the next required component -- \libpager.

If you load a stock\footnote{``Stock'' distributions are available at
  \href{https://cdimage.debian.org/cdimage/ports/latest/hurd-i386/}{\tt https://cdimage.debian.org/cdimage/ports/latest/hurd-i386/}}
Hurd distribution and run \netmsg\footnote{\netmsg is available at \href{https://github.com/BrentBaccala/hurd}{\tt https://github.com/BrentBaccala/hurd}} as shown above, you get a remote
file system that lets you read and write files, but executing remote
files doesn't work -- the program just hangs.
That's because Hurd, by default,\footnote{It's not too hard to change
  this behavior by changing the {\tt exec} server.} executes files
not by reading them into memory, but by {\tt mmap}'ing them into
memory.  This requires the {\it remote} file server's memory
management code to service {\it two} clients -- its local kernel, and
the remote one that is now trying to {\tt mmap} the file, and a stock
Hurd file server can't do that.  We need to upgrade {\tt libpager}.

\begin{comment}
For example, consider (again) what happens when node A drops its last
send right to a port on node B.  Node A sends a NO SENDERS message to
node B and forgets about the port.  It's possible that node B has sent
node A another send right for the same port, and that it's in transit
when node A sends the NO SENDERS message.  In that case, node A would
actually have an outstanding send right when node B receives the NO
SENDERS message, which would cause node B to incorrectly drop the send
right linked to node A.  I can see no way around this except a more
complicated handshake where the NO SENDERS message from node A to node
B would include a sequence number allowing node B to determine that it
hadn't sent any additional send rights, and a reply message from node
B to node A finalizing the port deallocation.  (what about SEQNOS?)
\end{comment}

@@

Hurd's implementation of the Mach memory management protocol is
implemented in a library called \libpager.  The original \libpager,
written in the mid-1990s, only supported a single client (i.e, a
single Mach kernel), and was thus suitable for single node operation.
Its API, however, is flexible enough to allow a more sophisticated,
multi-client \libpager as a drop-in replacement.  For dynamically
linked servers, all that is required is to update the system shared
library to obtain multi-node functionality.  Note, however, that
Hurd's root file system will almost always be implemented using a
statically linked server, since it needs to start quite early in
the boot process.

@@

How do we figure out which client is the local kernel?  One possible
solution is for \netmsg to listen on /servers/netmsg for an RPC that
accepts a send right and identifies which node currently hosts the
corresponding receive right.  \cite{interfaces} defined the {\tt
  norma_port_location_hint} RPC that might be useful.

@@

Two bits were allocated to each of two error codes.  {\tt ENOSPC} and
{\tt EDQUOT} were preserved; all other errors got collapsed into {\tt
  EIO}.  The remaining bits indicated if the page was in core, if we
were waiting for it to be paged in or out to the backing store, and if
the last attempt to write the data to the backing store had resulted
in an error.

@@

Handling error on unlock requests is somewhat tricky.
\cite{interfaces} states, in its {\bf NOTES} on {\bf
  memory_object_lock_request}:

@@

\section{Distributed Shared Memory}

Two shared memory implementations are available on modern Linux
systems -- the older System V shared memory API and the newer POSIX
API.  POSIX shared memory has not yet been implemented for Hurd, but
System V shared memory is implemented by mapping ``files'',
corresponding to shared memory keys, in the {\tt /dev/shm} portion of
the filesystem tree.  Linux uses a similar approach.  Thus, a
multi-client \libpager enables the use of inter-node shared memory.
What is required to make this happen is a shared {\tt /dev/shm} --
shared with full Mach IPC semantics, not just NFS -- so \netmsg is
required.

% The stock Hurd
% distribution currently (May 2019) implements {\tt /dev/shm} as a
% symlink to {\tt tmp}.

Replacing the root filesystem on Hurd requires a reboot, so unless
you've recompiled the system translators with the new \libpager, it's
better to mount a new filesystem on {\tt /dev/shm}, using the new
\libpager.  Ideally, this would done with Hurd's ramdisk translator,
         {\tt tmpfs}, but {\tt tmpfs} doesn't currently implement
the \libpager interface, so it's unusable for {\tt /dev/shm}.
For demonstration purposes, we can create an ext2 filesystem
and loopback mount it on {\tt /dev/shm}.

\begin{figure}[h]
\begin{mdframed}
\begin{verbatim}
root@qemu-hurd-1:~# dd if=/dev/zero of=shm bs=1M count=10    
10+0 records in
10+0 records out
10485760 bytes (10 MB, 10 MiB) copied, 0.1 s, 105 MB/s
root@qemu-hurd-1:~# mke2fs shm
mke2fs 1.44.6 (5-Mar-2019)
Creating filesystem with 2560 4k blocks and 2560 inodes

Allocating group tables: done                            
Writing inode tables: done                            
Writing superblocks and filesystem accounting information: done

root@qemu-hurd-1:~# settrans -ak /dev/shm ~/hurd-0.9.git20190331/build-deb/ext2fs/ext2fs.static shm
root@qemu-hurd-1:~# ls /dev/shm
lost+found
\end{verbatim}
\end{mdframed}
\end{figure}

This isn't ideal, because our shared memory is saved to disk, but it
serves to prove the functionality desired.  We can now start a \netmsg
server and run a program that allocates a shared memory segment,
writes a string of data into it, and waits for an acknowledgment
that it has been read:

\begin{figure}[h]
\begin{mdframed}
\begin{verbatim}
root@qemu-hurd-1:~# ~/github/netmsg/netmsg -s /dev/shm &
[1] 2533
root@qemu-hurd-1:~# ~/github/shm/Dave.Marshall/shm_server 
\end{verbatim}
\end{mdframed}
\end{figure}

On another machine, we use \netmsg to connect {\tt /dev/shm} to the
first machine.  The two machines now share their shared memory space,
and a client program is able to map the shared memory segment, read
the string, and acknowledge it.

\begin{figure}[h]
\begin{mdframed}
\begin{verbatim}
root@qemu-hurd-2:~# settrans -ak /dev/shm ./netmsg/netmsg/netmsg 192.168.55.101
root@qemu-hurd-2:~# ls /dev/shm
lost+found  sysvshm-162e
root@qemu-hurd-2:~# ~/shm_client 
abcdefghijklmnopqrstuvwxyz
root@qemu-hurd-2:~# 
\end{verbatim}
\end{mdframed}
\end{figure}

, but can also be implemented using the {\tt
  tmpfs} translator (a ramdisk), which is linked with \libpager.

unionfs can possibly be used to
overlay the various /run/shm's on each other.

{\tt mmap}, as currently implemented in Hurd's C library, only
allocates memory on the local node.  It does not seem difficult to
create a new translator that would service memory allocation requests
on remote nodes, in a manner similar to Linux's NUMA (Non-Uniform
Memory Access) support.  Linux's NUMA implementation uses a zonelist
for memory allocation.  Memory allocation requests are attempted
locally, and if this fails, then we fall back into the zonelist, which
gives us a list of other nodes to try for memory allocation.

\section{Distributed Tasks}

In the Hurd architecture, processes are localized to a node, but our
implementation of distributed shared memory allows cross-node forks
with full POSIX semantics.  A straightforward modification to Hurd's C
library would allow fork calls to spawn processes on other nodes, but
this has not yet been implemented.

Sharing threads between nodes is much more difficult, and seems
basically impossible to implement using Mach's current design.  The
main problem is Mach port handling.  Any thread in a task can listen
for messages on a Mach port, and messages are only delivered once, so
balancing the need to copy messages to multiple nodes (since any
thread could be the only one listening to a given port) while
guaranteeing unique delivery is quite problematic.

This suggests a natural division: processes are local to each node.  A
multi-threaded program can fork multiple times, and each forked
process can spawn multiple threads.  The processes are distributed
between nodes, while each process is local to a node and supports
multiple threads.  Since we have inter-node, inter-process shared
memory, almost all of the standard thread synchronization techniques
(semaphores, mutexes, condition variables) will work unmodified.  Just
as a typical multi-threaded process should try to avoid thread
contention on cache lines (typically 64 bytes), a multi-task process
should try to avoid task contention on memory pages (4 KB on the x86
architecture).

Since we have a working implementation of distributed shared virtual
memory, implementing cross-node forks doesn't seem too difficult.
There is a problem, however.  Mach allows the allocation of memory
pages unmanaged by any memory manager.  Unmanaged pages are typically
used for a process's routine data pages, but a cross-node fork would
require them to become managed in order to police multi-node access.

This suggests the need to attach a memory manager to the unmanaged
pages to allow them to be distributed across nodes, and this would
require an enhancement to the Mach kernel itself.  I think that
a new {\tt memory_object_attach} RPC would be used for this
purpose.

Once implemented, cross-node forks could be performed with full POSIX
semantics.  It's interesting that only at this point do we seem
to need a Mach kernel function that wasn't anticipated by Mach's
designers in the late 1980s.

\begin{comment}
\section{Blocking Behavior of {\tt mach_msg()}}

An issue with Mach's design caused me no little bit of grief.
Mach's {\tt mach_msg()} (as I read the documentation),
allows for non-blocking sends: according to \cite{interfaces}:

\end{comment}

\item Required Mach kernel enhancement

\item Optional Mach kernel enhancements

\begin{thebibliography}{9}

\bibitem{interfaces} {\it Mach 3 Kernel Interfaces}.

\bibitem{principles} {\it Mach 3 Kernel Principles}.

\end{thebibliography}

\appendix

\section{\libpager Pseudocode}

@@
